Demography and linked selection in wild and domesticated maize

### 2015-08-31 manuscript.tex
edits

\jri{this paragraph needs help} The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this hypothesis into question.}

\subsection{Demography of domestication}
\jri{section needs help} 
Although many other authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied  exclusively genic polymorphisms and have made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, making genic data inappropriate for estimating demography.
Because levels of diversity alone do not allow independent inference of the strength and duration of a bottleneck \cite{tenaillon2004selection}, previous investigations have instead estimated the ratio of these parameters.  
We have overcome this limitations through the use of the full joint SFS of maize and teosinte and consideration of a more realistic model of exponential growth, enabling us to estimate for the first time the strength of the domestication bottleneck.  
Given that modern maize exhibits $\approx 80\%$ of the diversity of teosinte, our estimate that  the effective population size of the initial maize  population was only $\approx 5\%$ of the ancestral teosinte population seems surprisingly low.
The impact of a strong domestication bottleneck is likely ameliorated to some degree by  maize-teosinte gene flow and the rapid post-domestication growth estimated in our model.

@@

Although we estimate that the modern effective size of maize is three times that of teosinte, the small size of our sample dramatically reduces our power to identify low frequency alleles most sensitive to rapid population growth \cite{keinan2012}.  
A better estimate of the effective size of modern maize comes from our analysis of singleton diversity in a large sample of maize landraces \cite{Hearne2015}.
These data suggest a modern effective size of nearly 1 million, nearly eight times higher than modern teosinte.
Nonetheless, missing data, sample size, and ascertainment bias in the genotyping data all bias this estimate downward as well.
We can, however, get reasonable estimates of the census size of modern maize.
There are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, with a lower limit of 25,000 individuals planter per hectare \cite{baden2001culture}.
Using this lower bound on planting density, ignoring all industrial maize agriculture, and even assuming the effective size is only 0.1\% of the census size, this still implies a modern effective population size of more than one billion.
Clearly, the effective size of modern maize is extremely large.

\subsection{Demography influences the efficiency of purifying selection}
\jri{section needs help} 
The observation that pairwise diversity in maize is less affected by the distance to the nearest gene than in teosinte is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

The reversal of this trend when we analyze only singleton diversity instead of pairwise diversity stands out as a notable. Every mutation begins as a singleton (an allele present in only one individual), and therefore singletons are, on average, the youngest class of alleles that can be observed. Unlike pairwise patterns of diversity, which are most heavily influenced by intermediate frequency alleles based on the definition of $\pi$ \cite{nei1979}, singleton diversity is most influenced by recent patterns of evolution. Hence, because our demographic estimation indicates dramatic expansion of maize $N_e$ in the recent past, we expect for purifying selection to presently operate more efficiently in maize than in teosinte. This observation is very clearly demonstrated by the fact that singleton diversity in maize is more impacted by the distance to the nearest gene than it is in teosinte, as was shown in Figure \ref{fig:purify}.

>>>>

\jri{paragraph needs work. i need to reread C-D}
The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this hypothesis into question.}

\subsection{Demography of domestication}
Although a number of  authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sties.  
By utilizing the full joint SFS, we are also able to, for the first time,  estimate the strength of the domestication bottleneck without requiring assumptions about its duration.  
Given that  maize landraces exhibit $\approx 80\%$ of the diversity of teosinte \citep{hufford2012}, our estimate of bottleneck of $\approx 5\%$ of the ancestral teosinte population seems surprisingly low.
The impact of a strong domestication bottleneck is likely ameliorated to some degree by both gene flow and the rapid post-domestication growth estimated in our model.

@@

Although we estimate that the modern effective size of maize is three times that of teosinte, the small size of our sample dramatically reduces our power to identify the low frequency alleles most sensitive to rapid population growth \cite{keinan2012}.  
A better estimate of the effective size of modern maize comes from our analysis of singleton diversity in a large sample of maize landraces \cite{Hearne2015}.
These data suggest a modern effective size nearly eight times higher than modern teosinte at $\approx 1$ million.
Even so, missing data, sample size, and ascertainment bias in the genotyping data likely bias this estimate downward as well.
An alternative approach is to use estimates of the census size of modern maize.
There are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, likely planted at a density of $\approx 25,000$ individuals  per hectare \cite{baden2001culture}.
Assuming the effective size is only 0.4\% of the census size (i.e. 1 female for every 1000 males), this still implies a modern effective population size of more than four billion.
While these rough estimates are likely quite poor, all of the evidence points to the fact that the effective size of modern maize is extremely large.

\subsection{Demography influences the efficiency of purifying selection}
\jri{section needs help} 
We find  a more pronounced decrease in nucleotide diversity around genes in teosinte compared to domesticated maize.
Because purifying selection is more efficient in larger populations  \cite{kimura1984},  this observation is consistent with our demographic model as well as previous work showing a smaller long-term effective population size in maize.  
This pattern of diversity is reversed, however, when we assess diversity at singleton polymorphisms.  
Under a simple neutral model, lower frequency polymorphisms will be younger and patterns of diversity of singleton polymorphisms should thus represent much more recent evolutionary history. 
We argue that the stronger decline in singleton diversity in and around genes in maize is the result of the increased efficacy of purifying selection  associated with the dramatic population expansion in maize following domestication.

This difference in recent $N_e$ should lead to an increased efficacy of purifying selection in the recent past of  Differences in diversity observed at  demographic estimation indicates dramatic the recent past, we expect for purifying selection to presently operate more efficiently in maize than in teosinte. This observation is very clearly demonstrated by the fact that singleton diversity in maize is more impacted by the distance to the nearest gene than it is in teosinte, as was shown in Figure \ref{fig:purify}.
